---
title: "Explore Gapminder with dplyr and ggplot2"
author: "Lisa Wei"
date: '2017-09-23'
output: github_document
---


This is an exploration of the Gapminder dataset using tools and functions from the dplyr package.

## Bring rectangular data in
```{r setup, message=FALSE}
## load the gapminder package
#install.packages("gapminder")
library(gapminder)
## load the tidyverse package (containing dplyr)
library(tidyverse)

#load other useful packages
library(ggExtra)
library(knitr)
library(kableExtra)
```

### "Smell" test the data

Q1.1: Is it a data.frame, a matrix, a vector, a list?
A: Gapminder is a tibble, data.frame, and a list.

```{r}
class(gapminder)
```


Q1.2: What’s its class?
A: Gapminder is a list as indicated by `typeof()`. But it is also a tibble and data frame if one uses `class()`.

```{r}
typeof(gapminder)
```

Q1.3: How many rows/observations?
A: Number of rows:
```{r}
nrow(gapminder); 
```

Q1.4: How many columns/variables?
A: number of columns

```{r}
ncol(gapminder)
```


Double check with dim() showing dimensions of the data
```{r}
dim(gapminder); 
```

Q1.5: Can you get these facts about “extent” or “size” in more than one way? Can you imagine different functions being useful in different contexts?

A: Can just type in gapminder to get a sneak peak of the data as well as the dimensions. Also, other functions that could be useful is using `str()` which shows the overall dimension and type of the dataset and the type of each variable/column. The `str()` function is very useful is one wants a quick way to access the class of each variable in order to do operations on certain columns.
 

```{r}

gapminder

str(gapminder)
```


Q1.6: What data type is each variable?
A: Can be answered with `str()` as seen above: 

- Country is factor   
- Continent is factor   
- Year is integer     
- LifeExp is number     
- Pop is integer    

Can also check the data type of each variable one by one:
```{r}
class(gapminder$country) 
class(gapminder$continent)
class(gapminder$year)
class(gapminder$lifeExp)
class(gapminder$pop)
class(gapminder$gdpPercap)
```


## Explore individual variables
Explore categorical variables `continent` and `country`. Explore quantitative variables `population` and `lifeExp`.   

Q2.1 What are possible values (or range, whichever is appropriate) of each variable?

A: For continent and country, since they categorical, below are all possible values for each variable. There are in total 142 distinct countries and 5 distinct continents in the gapminder dataset.


```{r}
unique(gapminder$continent)
unique(gapminder$country)

n_distinct(gapminder$continent)
n_distinct(gapminder$country)
```

For population and lifeExp, since they are quantitative, let's explore the the range. As seen below, the range of pop is `r min(gapminder$pop)` to `r max(gapminder$pop)`. The range of lifeExp is `r min(gapminder$lifeExp)` to `r max(gapminder$lifeExp)`.

```{r}
range(gapminder$pop)
range(gapminder$lifeExp)
# can also check with min and max
min(gapminder$pop)
max(gapminder$pop)

min(gapminder$lifeExp)
max(gapminder$lifeExp)
```


Q2.2: What values are typical? What's the spread? What's the distribution?

A: For continent, Africa shows up with the highest frequency. For country, all countries show up at equal frequency of 12 times. Since all countries have the same frequency, I will only plot a bar graph for continent.

```{r}
summary(gapminder)
## sorting from highest to lowest frequency for continents and countries
sort(table(gapminder$continent), decreasing=TRUE)
sort(table(gapminder$country), decreasing=TRUE)


## a bar plot to visualize the frequencies
ggplot(data=gapminder, aes(continent)) + geom_bar() + ggtitle("Frequency of Each Continent in the Gapminder Dataset") 


```


```{r}
summary(gapminder)
```

For population, as shown in the summary table just above, the minimum is `r as.integer(min(gapminder$pop))` and maximum is `r as.integer(max(gapminder$pop))`. Mean = `r as.integer(mean(gapminder$pop))` and sd = `r as.integer(sd(gapminder$pop))`. For life expectancy, the minimum is `r as.integer(min(gapminder$lifeExp))` and maximum is `r as.integer(max(gapminder$lifeExp))`. Mean = `r as.integer(mean(gapminder$lifeExp))` and sd = `r as.integer(sd(gapminder$lifeExp))`.

Below are histograms with additional features:

1. A gradient of the colour scheme that varies according to the frequency of the values.       

2. A normal distribution curve to compare the distribution of the population/life expectancy data to a theoretical model of a normal curve: The curve does not seem to fit the data well at all, as indicated by the red flat line in each plot.

```{r, message=FALSE, echo=FALSE}
ggplot(data=gapminder, aes(pop)) + 
  geom_histogram(aes(fill=..count..), binwidth=5e+07) + 
     scale_y_continuous(name = "Count") +
        stat_function(fun = dnorm, colour = "red",args = list(mean = mean(gapminder$pop, na.rm = TRUE),
                                 sd = sd(gapminder$pop, na.rm = TRUE))) +
          annotate("text", x = 5e+08, y = 1000, label = "mean = 29601212, sd=106157897") +
            ggtitle("Distribution of the population in gapminder") + xlab("Population")

ggplot(data=gapminder, aes(lifeExp)) + 
  geom_histogram(aes(fill=..count..)) + 
     scale_y_continuous(name = "Count") +
        stat_function(fun = dnorm, colour = "red",args = list(mean = mean(gapminder$lifeExp, na.rm = TRUE),
                                 sd = sd(gapminder$lifeExp, na.rm = TRUE))) +
          annotate("text", x=40, y=150, label = "mean = 59, sd=12") +
            ggtitle("Distribution of the life expectancy in gapminder") + xlab("Life Expectancy")


```


## Explore various plot types and practicing using `filter()`, `select()`, and `%>%`

- Scatterplot of life expectancy vs GDP per capita, separate plot for each continent and size of the points indicating year. Added a line of best fit to the data.

```{r, message=FALSE}
p <- ggplot(gapminder, aes(x=gdpPercap, y=lifeExp)) + 
        geom_point(alpha = (1/3), size = 3) + facet_wrap(~ continent) +
          geom_smooth(lwd = 1.5, se = FALSE) + 
            facet_wrap(~continent)  + ggtitle("Life expectancy vs gdpPercap for each continent") + xlab("GDP per capita") + ylab("Life expectancy")

p
```

- Violin plots of population by year after 2000, grouped by 2 continents Africa and Amercias (compare 2 continents). Also I'm piping filtered data into ggplot in order to practice using filter, select, and piping into ggplot.

```{r, message=FALSE}
filter(gapminder, (continent=="Africa" & year>2000) | (continent=="Americas" & year>2000)) %>% ggplot(aes(x=factor(year), y=gdpPercap)) + geom_violin(aes(fill=factor(continent))) + ggtitle("Distribution of population for year>2000 in Africa vs Americas") + xlab("Year") + ylab("GDP per capita")
```



- Histogram of life expectancies, between 2 different continents on the same plot.

```{r, message=FALSE}
filter(gapminder, continent %in% c("Africa","Americas")) %>% select(lifeExp,continent) %>% ggplot(aes(x=lifeExp, fill=factor(continent))) + 
    geom_histogram(alpha = 0.5, position="identity") + ggtitle("Distribution of life expectacies between Africa and America") + xlab("Life expectancy") + ylab("Frequency")

```

###Additional practices: Practicing using arrange() and mutate()

First, arrange the data by year, and then sort by lifeExp in decreasing order.

```{r}
arranged_dat <- arrange(gapminder, year, desc(lifeExp)) %>%
  select(year, lifeExp, everything())
```


Second, calculate the total gross domestic product by multiplying gdpPercap by the population and add that to the gapminder data as a separate column and visualize this in a plot. The size of the points corresponds to the size of the population. I also added a line of best fit.

```{r, message=FALSE}
g <- arranged_dat %>% mutate(gdp = gdpPercap * pop) %>% ggplot(aes(lifeExp, gdp)) + 
  geom_point(aes(size = pop), alpha=0.25, colour="black") + 
   geom_smooth(method="lm", se=F) + ggtitle("Total GDP vs life expectancy") + xlab("GDP") + ylab("Life Expectancy")
g
```

Make a boxplot of gdp per billion comparing between America, Americas, Asia over time after year 2000.

```{r, message=FALSE}
g <- mutate(gapminder, gdp = gdpPercap * pop, gdpBill = round(gdp/1000000000, 1)) %>% filter((continent %in% c("Africa","Americas","Asia")) & (year>2000)) %>%    
        ggplot(aes(x=factor(year),y=gdpBill,fill=factor(continent))) +
          geom_boxplot() +
            xlab("year") + ylab("life expectancy") +
              ggtitle("Distribution of GDP per billion for Africa, Americas and Asia after year 2000")

g
```


## Assessing the line of code shown below. Present numerical tables using `kable()`. Exploring more `dplyr` functions.

> filter(gapminder, country == c("Rwanda", "Afghanistan"))

This code is wrong as seen below. It seems that the above line of code outputs only combinations of all variables within the data where the year didn't repeat for rows containing Rwanda and Afghanistan, after filtering for the selected countries. To get all the data where the country is either Rwanda and Afghanistan, the `==` should be changed to `%in%` as shown below:

> filter(gapminder, country %in% c("Rwanda", "Afghanistan"))

Let's compare the two lines of code to see which works properly:
```{r}
t1 <- filter(gapminder, country == c("Rwanda", "Afghanistan"))
dim(t1)
t1

t2 <- filter(gapminder, country %in% c("Rwanda", "Afghanistan"))
dim(t2)
t2

identical(t1,t2)

```

The second one included more data than the first version of the code, and these additional rows seem to contain data within Rwanda and Afghanistan, so the second version of the code must be the correct one.

Now, visualize in a nice table:

```{r}
filter(gapminder, country %in% c("Rwanda", "Afghanistan")) %>%
  knitr::kable(format="markdown",align=c(rep('c',times=7)), padding=2, col.names=c("Country","Continent","Year","Life Expectancy","Population","GDP per Capita"))
```


## Additional experimentation with dplyr operations on gapminder

- Counting up data

```{r}
gap_filtered <- filter(gapminder, continent %in% c("Africa","Americas"))
group_by(gap_filtered, continent) %>%
  summarize(n = n()) %>%          
     knitr::kable(format="markdown",align=c(rep('c',times=7)), padding=2, col.names=c("Continent","Count"))
```


- Calculating the average life expectancy for each country

```{r}
group_by(gap_filtered, continent) %>%
  summarize(avg_lifeExp = mean(lifeExp)) %>%
    knitr::kable(format="markdown",align=c(rep('c',times=7)), padding=2, col.names=c("Continent","Average Life Expectancy"))

```


## Report your process
- What was hard/easy.

It was good that we were working with a dataset (Gapminder) that was already introduced in class since I already had some familiarity of the variables and general statistics. This made it easier to think about what types of plots would be useful and suitable for exploring/representation the Gapminder dataset.
The hard part was figuring out how to draw the various types of plots with ggplot2. Specifically, it took me a while to figure out how to group the violin and boxplots by a specific categorical variable. Also, figuring out where to include the `aes()` specifications and the colours to use in the plot to draw points or bars were time-consuming. I basically Googled every question just as Vincenzo suggested in class.

- Problems you solved

1. Aligning columns in the output table by the `kable()` function. This was achieved using `align=c(rep('c', times=7)`. I found this in a Stackoverlow thread as indicated below.

2. How to overlay histograms (in my case distributions of life epectancy for 2 continents) onto one plot. This was achieved by usig `fill=factor(continent)` in the global `ggplot()` call.

3. Figuring out what was wrong, if there was anything wrong, with the line of code (shown below) was tricky.

> filter(gapminder, country == c("Rwanda", "Afghanistan"))

I had assumed that the above line of code wouldn't output anything since there are no countries that has a name `c("Rwanda","Afghanistan")`. But when I actually ran the code, it gave me an output. And it was only until I ran what I thought was the correct version using the `%in%` function that I realized the first version didn't output as much data as my version of the code. So always explicitly check, and never assume anything!


- Helpful tutorials you read:
1. [plotting distributions in ggplot2](http://www.cookbook-r.com/Graphs/Plotting_distributions_(ggplot2)/) for plotting 2 separate distributions on the same plot.

2. [stat_function() documentation](https://www.rdocumentation.org/packages/ggplot2/versions/0.9.0/topics/stat_function) which I referenced when I wanted to draw the normal distribution curves and overlay that on top of the histograms.    

3. [Making histograms where the color scheme changes with the frequency of the values](http://t-redactyl.io/blog/2016/02/creating-plots-in-r-using-ggplot2-part-7-histograms.html). This functionality was super neat and really helped visually with highlighting the parts of the histogram with high frequency.

4. [dplyr functions on a single dataset by Jenny Bryan](http://stat545.com/block010_dplyr-end-single-table.html). I referenced this for many additional functions such as `group_by()`, `summarize()`, `mutate()`, and `arrange()`.

5. [Aligning columns with kable](https://stackoverflow.com/questions/41365502/aligning-columns-with-knitr-kable-function). This Stackoverflow thread was where I got the answer to how to align my columns and format the table nicely.


