---
title: "API"
author: "Lisa Wei"
date: '2017-11-30'
output: github_document
---

For this assignment, I've decided to use the OMDb API: `https://www.omdbapi.com/` to get info about movies.

### Need a key to access the information: ask for an API key
I requested an API key: `apikey=df5f9e85`

#### Load the relevant packages

```{r setup}
library(httr)
#library(XML)
library(dplyr)
library(glue)
library(tibble)
library(knitr)
```


## Example of getting movie info from the web

```{r setup2}
thor_result <- httr::GET(url = "http://www.omdbapi.com/?t=Thor&y=2017&apikey=df5f9e85")
 
thor_content <- content(thor_result)
#thor_content %>% View()
```


## Let's write a function to retrieve movie info 

```{r function}
get_movie_TY <- function(title, year){
	query_string <- glue("http://www.omdbapi.com/?t={title}&y={year}&apikey=df5f9e85")
	movie_result <- GET(query_string)
	movie_content <- content(movie_result)
	return(movie_content)
}
		
```


## Getting movie that contains `babe` or `time` in title and also `The Matrix`

```{r analysis}
babe <- get_movie_TY(title="babe",year="")

neo <- get_movie_TY(title="Matrix",year="1999")

time <- get_movie_TY(title="time",year="")
```




## Formatting the outputs into table format

The tibble package has a function enframe() that solves this problem by coercing nested list objects to nested tibble ("tidy" data frame) objects. 


```{r dataframe}
neo_df <- t(enframe(neo)); colnames(neo_df) <- names(neo); neo_df <- neo_df[-which(rownames(neo_df)=="name"),]
babe_df <- t(enframe(babe)); colnames(babe_df) <- names(babe); babe_df <- babe_df[-which(rownames(babe_df)=="name"),]
time_df <- t(enframe(time)); colnames(time_df) <- names(time); time_df <- time_df[-which(rownames(time_df)=="name"),]

df <- rbind(neo_df, babe_df, time_df); colnames(df) <- names(neo)

```


## Final output table with all the movies/info extracted from the web

```{r table_output}
kable(df)

```







